Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-03-26 Paper/manuscript_v2/manuscript_07_01.tex
Minor edits and to dos in sec 4

The accuracy and speed of the HSGP model depends on several interrelated factors, most notably on the number of basis functions and on the boundary condition of the Laplace eigenfunctions. Furthermore, appropriate values for these factors will depend on the non-linearity of the estimated function, which is in turn characterized by the lengthscale of the covariance function.
In this section, we analyze the effects of the number of basis functions and the boundary condition on the approximation accuracy. We present recommendations on how they should be chosen and diagnostics to check the accuracy of the obtained approximation.

Ultimately, these recommendations are based on the relationships among the number of basis functions, the boundary factor and the lengthscale of the function, which depend on the particular choice of the kernel function. In this work we built these relationships for the square exponential covariance function and Matern ($\nu$=3/2) covariance function in the present section, and for the periodic squared exponential covariance function in Appendix \ref{ch5_sec_periodic}. For other kernels, the relationships will be slightly different, in function of mainly the smoothness or wigglyness of the kernel effects.

@@

As explained in Section \ref{ch5_sec_method}, the approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, for instance in a one-dimensional input space $\Omega=[-L,L]\subset {\rm I\!R}$:
%
\begin{equation*}
k(\tau) = \sum_{j=1}^{\infty} S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0), 
\end{equation*}

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}) for the unidimensional case and in equations (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{equation}\label{eq_diff_covs}
\Big( \int k(\tau) \mathrm{d}\tau - 
\int \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0) \mathrm{d}\tau \Big) < \varepsilon.
\end{equation}

The finite number $m$ of basis functions in the approximation needed to satisfy equation (\ref{eq_diff_covs}) depends on the non-linearity of the function to be learned, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see equations (\ref{eq_eigenvalue}), (\ref{eq_eigenfunction}), (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \Omega$. We now define $L$ as
%
\begin{equation}\label{eq_boundary}
L=c \cdot S,
\end{equation}

\noindent where $S$ (for $S > 0$) represents the half-range of the input space, and $c$ (for $c \geq 1$) is the proportional extension factor. In the following, we will refer to $c$ as the boundary factor of the approximation. The boundary factor can also be regarded as the boundary $L$ normalized by the half-range $S$ of the input space.

We start with an illustration on how the number of basis functions $m$ and boundary factor $c$ influences the accuracy of the HSGP approximations, separately. For this purpose, a set of noisy observations are drawn from an exact GP model with lengthscale $\ell=0.3$ and marginal variance $\alpha=1$ of the kernel function, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $L$ are fitted to this data. In this example, the lengthscale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictions of the estimated function and on the covariance function itself. For $c$ fixed to a large enough value, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation and the non-linearity of the estimated function, in the sense that fewer basis functions inaccurately imply larger lengthscales and consequently more linear functional forms.
The higher the "wigglyness" of the function to be estimated, the more basis functions will be required. If $m$ fixed to a large enough value, Figure \ref{fig2_Post_L} shows that $c$ mainly affects the approximation near the boundaries as well as covariances at long distances.

>>>>

The accuracy and speed of the HSGP model depends on several interrelated factors, most notably on the number of basis functions and on the boundary condition of the Laplace eigenfunctions. Furthermore, appropriate values for these factors will depend on the degree of non-linearity of the estimated function, which is in turn characterized by the lengthscale of the covariance function.
In this section, we analyze the effects of the number of basis functions and the boundary condition on the approximation accuracy. We present recommendations on how they should be chosen and diagnostics to check the accuracy of the obtained approximation.

Ultimately, these recommendations are based on the relationships among the number of basis functions, the boundary condition and the lengthscale of the function, which depend on the particular choice of the kernel function. In this work we investigate these relationships for the square exponential covariance function and Mat{\'e}rn ($\nu$=3/2) covariance function in the present section, and for the periodic squared exponential covariance function in Appendix \ref{ch5_sec_periodic}. For other kernels, the relationships will be slightly different, in function of mainly the smoothness or wigglyness of the kernel effects. \todo{The last sentence is not very clear. }

@@

As explained in Section \ref{ch5_sec_method}, the approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, for instance in a one-dimensional input space $\Omega=[-L,L]\subset {\rm I\!R}$
%
\begin{equation*}
k(\tau) = \sum_{j=1}^{\infty} S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0), 
\end{equation*}

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}) for the unidimensional case and in equations (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$
%
\begin{equation}\label{eq_diff_covs}
\Big( \int k(\tau) \mathrm{d}\tau - 
\int \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0) \mathrm{d}\tau \Big) < \varepsilon.
\end{equation}

\todo{I think the equation above is wrong, it should compared the areas under the curves and not the difference between the exact and approximate covariance function. It should the total variation difference between the true kernel and the approximate kernel, i.e. something like
\begin{align*}
\int|k(\tau) -S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0) \mathrm{d}\tau |\mathrm{d}\tau 
\end{align*}.
How did you compute it in the code?
}

The specific numbe of basis functions  $m$ in the approximation needed to satisfy equation (\ref{eq_diff_covs}) depends on the degree of non-linearity of the function to be estimated, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see equations (\ref{eq_eigenvalue}), (\ref{eq_eigenfunction}), (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi = \max_{i}|x_i|$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \Omega$. We now define $L$ as
%
\begin{equation}\label{eq_boundary}
L=c \cdot S,
\end{equation}

\noindent where $S$ (for $S > 0$) represents the half-range of the input space, and $c \geq 1$ is the proportional extension factor. In the following, we will refer to $c$ as the boundary factor of the approximation. The boundary factor can also be regarded as the boundary $L$ normalized by the half-range $S$ of the input space.

We start by illustrating how the number of basis functions $m$ and boundary factor $c$ influence the accuracy of the HSGP approximations individually. For this purpose, a set of noisy observations are drawn from an exact GP model with lengthscale $\ell=0.3$ and marginal variance $\alpha=1$ of the covariance function \todo{which covarariance function?}, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $L$ are fitted to this data. In this example, the lengthscale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictions of the estimated function and on the covariance function itself. For $c$ fixed to a large enough value, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation and the non-linearity of the estimated function, in the sense that fewer basis functions inaccurately imply larger lengthscales and consequently more linear functional forms.
The higher the "wigglyness" of the function to be estimated, the more basis functions will be required. If $m$ fixed to a large enough value, Figure \ref{fig2_Post_L} shows that $c$ mainly affects the approximation near the boundaries.
